### 2020-08-03 paper/sex_selection_spacing-ver4.tex
Combine and reorganize birth interval sections
- Make one section
- Discuss results w/o sex selection
- Discuss results w sex selection

% Explain question and what I am doing
The first question I address is how birth spacing responded to the significant changes, 
especially the spread of sex selection, that India experienced. 
I estimate the competing risk hazard model for each spell, education, and period 
combination and then predict 25th, 50th, and 75th percentiles birth intervals in 
months, the sex ratio, and the probability of having a birth in that spell.
Sex selection increases all three percentile birth intervals, but the effects of abortions 
are easiest to observe in the 75th percentile birth interval, 
and it is also the changes in the longest birth intervals that will likely the largest 
impact on fertility measures.
The 25th percentile captures changes in the short spacing, which is the most likely to 
have an adverse impact on health.

Figures \ref{fig:spacing_low_rural} through \ref{fig:spacing_highest_urban}
show the predicted outcomes for the four education levels by area of residence.
The underlying values with bootstrapped standard errors are available in the online 
appendix.
The sex ratio graphs also show the natural sex ratio, approximately 51.2 percent 
\citep{ben-porath76b,jacobsen99,Portner2015b}.
Results for urban women without education, rural women with twelve or more years of 
education, and the fourth spell for women with twelve or more years of education are
presented in the online appendix because they are based on relatively few observations.

@@

The results show three broad trends.
% [Birth intervals]
First, 
over the four decades, birth intervals have lengthen, whether measured by the
25th, 50th, or 75th percentile.
These increases are larger the higher the education of the mother, the higher the parity, 
and the higher the percentile measure.

% [Parity progression probability]
Second, 
the likelihood of a next birth has fallen over time, and the declines are
larger the higher the education, the higher the parity, and with the presence of at
least one son. 
Within a given spell and period, parity progressions are lower in urban than in rural areas, 
if at least one son is present, and the more educated the mother.

% [Sex ratios]
Finally, 
the sex ratios of next births has become more male dominated for women with no
sons, while there are no clear trends for the other sex compositions.
The increases in the predicted percent boys are larger the higher the education and the 
higher the parity.
Furthermore, for a given spell and period, sex ratios in the absence of a son are higher 
in urban than in rural areas, the higher the education of the mother, and the higher
the parity.
There is some evidence of unnaturally high percentage boys born even when women have one 
son, although most are clearly lower than in the absence of a son and the failing fertility
make more of those estimate noisy.

% [25th percentile birth intervals]
Despite the increases in birth intervals and the decline in fertility, a remarkably high 
proportion of birth intervals are still very short.
For all but the most educated, more than 25 percent of women have their second and their
third child within 24 months of their previous child.
Furthermore, even with the larger increases in birth intervals with higher parity, the 
25th percentile birth intervals for the fourth spell is around 24 months for women with 
less than eight years of education.

% [median birth intervals]
The median birth intervals have increased more than the 25th percentile birth intervals
but are still also relatively short.
Although most median birth intervals have increased by three to six months over the four 
decades, the majority of median birth intervals are still at 36 months or below, 
with the shortest only 29 months.
For comparison, fast changing countries have median birth intervals that increase around 
3.5 months per decade with declining fertility \citep{Rutstein2011,Casterline2016}.
Furthermore, the NFHS reports show median closed birth intervals that have barely
moved over time, staying at approximately 31 months, underscoring the importance of 
accounting for censoring of birth spells when trying to understand how birth spacing 
changes over time.

An important exception to the relatively short median intervals come from the better 
educated women, especially in the absence of sons.
The longest median interval was 48 months for the third spell for urban women with twelve 
or more years of education who had no boys, an increase of almost 15 months over
four decades.
Similary, urban women with eight to eleven years of education and no sons had median
birth intervals of around 40 months, an increase of about 10 months.
Finally, the best educated women now have a median second birth interval of around 42
months whether their first child was a boy or a girl.

% [75th percentile birth intervals]
The most substantial differences across women come from the longer birth intervals, 
captured by the 75th percentile birth intervals.
For the two lowest education groups the increases in the 75th percentile birth 
intervals are relatively small for the second spell and slightly more substantial for the 
third spell at around five months.
However, the fourth spell show significant increases, especially for women with two
or three sons, with increases of fifteen to twenty months.
The fourth spell is also where most of the divergence across sex compositions happen,
but even for women with no or one son, the 75th percentile birth intervals have
increased by a half to a full year.
For women with eight to eleven years of education, the 75th percentile spacing length for 
women with two girls in the third spell increased by more than 15 months, mainly as the 
result of more widespread use of sex selection as the sex ratios above indicate.

% [Exceptions to the results]
The best-educated women are an especially interesting case that run somewhat counter
to the results above.
First, despite a rapidly declining probability of having a third child when they already 
have a son, there is little change in the associated birth intervals and all are lower for 
the third spell than for the second.
Second, the massive increase in the sex ratio to almost 80 percent boys for women with two 
girls is associated with a more than 20 month increase in the 75th percentile birth 
interval.
The third spell is also the case with the clearest reversal in the spacing pattern by sex 
composition; the birth intervals with two girls are consistently longer than with one
or two sons no matter the percentile used.

What do these results tell us about the role of son preference and the extent to which 
the introduction of sex selection changed birth spacing across sex compositions?

The clearest illustration of how the combination of son preference and the introduction of 
sex selection impacted birth spacing comes from the third spell of the best-educated urban 
women.
Women with two girls have a predicted sex ratio for their third child approaching 80 
percent boys and a 75th percentile birth interval of almost 70 months, about 13 months
longer than if they had at least one son.
Even more striking is the change for this group when sex selection was first introduced.
The 75th percentile birth interval with two girls increased from 48 months to 
64 months with two girls, while the other sex compositions showed a 
slightly decrease from around 55 months to 54 months.
Theses changes may even be an underestimate of the changes in birth spacing sine
this particular group appear to have had access to sex selection even before it became
widespread as shown by the unequal sex ratio for the 1972--1984 period for women with
two girls.

Even for the women the least likely to use sex selection, such as rural women with no 
education, son preference shows up clearly. 
For this group, the shortest spacing---across all spells---is when they have only girls. 
Notably, the difference in birth intervals across sex compositions has grown over 
time as spacing when sons are present increased.
Despite the strong son preference in spacing and sex ratios, there is also some indication 
that parents prefer a mix of boys and girls;
women with two boys \emph{and} a girl consistently have the lowest parity progression 
probability in the fourth spell across education levels.

Did the predictions of a turn-around in the use of sex selection come true?
There is no clear evidence for or against a reversal in the use of sex selection, with 
some showing increases in sex ratios between the last two periods, others little change, 
and some a decline.
The best-educated women is again a good illustration.
The percent boys born to women with two girls continued to increase over the last two
periods, but the likelihood of a third birth declined and when the first child was a girl, 
the sex ratio for the second birth dropped slightly as did the probability of a second birth. 
There is, however, also cases where there is clearly no abatement in the increasing use of 
sex selection. 
The sex ratios in the absence of girls continued to increase for rural women with
one to seven years of education and the likelihood of an additional birth remained high.

% [Interpretation]
How do these changes in birth intervals tie in with the changes in India over the four
decades?
% [TK economic growth? -> higher return to education for males -> lower fertility and
% higher investment in boys' education -- has anybody looked at this?]
First, the parity progression probabilities have been falling over time across all 
education levels, especially for women with at least one son.
Falling fertility is associated with increases in longer spacing, although it is still an 
unresolved question why \citep{Casterline2016}.
What is especially interesting is that the results for the most educated women show that 
the rapidly declining probability of a third birth for women with one or two sons did not 
lengthen the birth intervals at any of the percentiles.
One possibility is that the most educated women have both better access to modern
contraceptive and are better at using traditional contraceptive methods 
\citep{Rosenzweig1989}.

Second, there is evidence that in situations where sex selection does not play a role 
in birth intervals, women with the smallest change in birth intervals are also those most 
likely to work.
For example, the increases in average third spell birth intervals for urban women with
one boy and one girl over the four decades were 2.7, 3.4, 5.8, and 1.8 months from lowest 
to highest education groups.%
\footnote{
For women with two sons, the numbers were 4.3, 6.3, 7.0, and 3.0, respectively.
See the online appendix tables for the average birth intervals.
}
This means that the group with the lowest labor force participation, women with eight to 
eleven years of education, was also the group with the largest increases in average 
spacing, likely driven by the substantial improvement in household income that came
from economic growth.

@@

>>>>

% Explain question and what I am doing
The first question I address is how birth spacing responded to the significant 
changes in India.
Figures \ref{fig:spacing_low_rural} through \ref{fig:spacing_highest_urban}
show the 25th, 50th, and 75th percentiles birth intervals in 
months, the sex ratio, and the probability of having a birth for each spell
for the four education levels by area of residence.%
\footnote{
Results for urban women without education, rural women with twelve or more years of 
education, and the fourth spell for women with twelve or more years of education are
presented in the online appendix because they are based on relatively few observations.
}
The underlying values with bootstrapped standard errors are available in the online 
appendix.
The sex ratio graphs also show the natural sex ratio, approximately 51.2 percent 
\citep{ben-porath76b,jacobsen99,Portner2015b}.

% [not sure where this goes or simply delete]
%Sex selection increases all three percentile birth intervals, but the effects of abortions 
%are easiest to observe in the 75th percentile birth interval, 
%and it is also the changes in the longest birth intervals that will likely the largest 
%impact on fertility measures.
%The 25th percentile captures changes in the short spacing, which is the most likely to 
%have an adverse impact on health.

@@

%The results show three broad trends.
%% [Birth intervals]
%First, 
%over the four decades, birth intervals have lengthen, whether measured by the
%25th, 50th, or 75th percentile.
%These increases are larger the higher the education of the mother, the higher the parity, 
%and the higher the percentile measure.

% [Parity progression probability]
The parity progression and the sex ratio show two broad trends.
First, 
in line with the observed fall in the total fertility rate, the likelihood of a next birth 
has fallen over time.
The declines in the likelihood of a next birth are larger the higher the education, 
the higher the parity, and with the presence of at least one son. 
Within a given spell and period, parity progressions are lower in urban than in rural areas, 
if at least one son is present, and the more educated the mother.

% [Sex ratios]
Second, 
the spread of sex selection shows clearly in the sex ratios of next births, which has 
become more male dominated for women with no sons.
The increases in the predicted percent boys are larger the higher the education and the 
higher the parity.
There are no clear trends for the other sex compositions.
Furthermore, for a given spell and period, sex ratios in the absence of a son are higher 
in urban than in rural areas, the higher the education of the mother, and the higher
the parity.
There is some evidence of unnaturally high percentage boys for some women with one 
son, although most are clearly lower than in the absence of a son and the failing fertility
make these estimate noisy.

% [Birth intervals w/o sex selecition]

To separate out the effects of the introduction of sex selection and the other changes in
India, I first discuss how birth intervals have changed in situations where sex selection are 
less likely to be used.
This groups covers women with no education, no matter the sex composition of their children, 
and women with any education who has one or two sons already. 
[This is not a hard and fast separation ]

% [median birth intervals]
Most median birth intervals have increased by three to six months over the four 
decades, but the majority of median birth intervals are still at 36 months or below, 
with the shortest only 29 months.%
\footnote{
The NFHS reports show median closed birth intervals of approximately 31 months, which have 
barely moved over time, underscoring the importance of accounting for censoring when examining 
birth spacing.
}
These changes are substantially smaller than the 3.5 months increase per decade found
in other countries with declining fertility \citep{Rutstein2011,Casterline2016}.

% [25th percentile birth intervals]
Despite the increases in median birth intervals and the decline in fertility, a remarkably 
high proportion of birth intervals are still very short.
For all but the most educated, more than 25 percent of women have their second and their
third child within 24 months of their previous child, which is substantially below
the 24 between \emph{pregnancies} recommended by the WHO.
Furthermore, even with the larger increases in birth intervals with higher parity, the 
25th percentile birth intervals for the fourth spell is around 24 months for women with 
less than eight years of education.

% [75th percentile birth intervals]
The most substantial differences across women come from the longer birth intervals, 
captured by the 75th percentile birth intervals.
For the two lowest education groups the increases in the 75th percentile birth 
intervals are relatively small for the second spell and slightly more substantial for the 
third spell at around five months.
However, the fourth spell show significant increases, especially for women with two
or three sons, with increases of fifteen to twenty months.

These results are in line with prior research showing that falling fertility is 
associated with increases in longer spacing, although it is still an 
unresolved question why \citep{Casterline2016}.
The parity progression probabilities have been falling over time across all 
education levels, especially for women with at least one son, and have been
failing less for the less educated women.

However, despite the rapidly declining probability of a third birth for the best-educated
women when they already have a son, the birth intervals changed little for any of the percentiles,
and all are lower for the third spell than for the second.
One possibility is that the most educated women have both better access to modern
contraceptive and are better at using traditional contraceptive methods 
\citep{Rosenzweig1989}.

% [Interpretation]
% How do these changes in birth intervals tie in with the changes in India over the four
% decades?
% [TK economic growth? -> higher return to education for males -> lower fertility and
% higher investment in boys' education -- has anybody looked at this?]
% First, the parity progression probabilities have been falling over time across all 
% education levels, especially for women with at least one son.

Women with the largest increases in birth intervals are also those least likely to work.
For example, the increases in average third spell birth intervals for urban women with
one boy and one girl over the four decades were 2.7, 3.4, 5.8, and 1.8 months from lowest 
to highest education groups.%
\footnote{
For women with two sons, the numbers were 4.3, 6.3, 7.0, and 3.0, respectively.
See the online appendix tables for the average birth intervals.
}
This means that the group with the lowest labor force participation, women with eight to 
eleven years of education, was also the group with the largest increases in average 
spacing, likely driven by the substantial improvement in household income that came
from economic growth.

Son preference still show up clearly, even for the women the least likely to use sex 
selection, such as rural women with no education. 
For this group, the shortest spacing---across all spells---is when they have only girls. 
Notably, the difference in birth intervals across sex compositions has grown over 
time as spacing when sons are present increased.

% [Birth intervals with sex selection]

The clearest illustration of how the combination of son preference and the introduction of 
sex selection affected birth spacing comes from the third spell of the best-educated urban 
women.
Women with two girls have a predicted sex ratio for their third child approaching 80 
percent boys and a 75th percentile birth interval of almost 70 months, about 13 months
longer than if they had at least one son, and an increase of more than 20 months over
the four decades.

Even more striking is the change for this group when sex selection was first introduced.
The 75th percentile birth interval with two girls increased from 48 months to 
64 months with two girls in a decade, while the other sex compositions showed a 
slightly decrease from around 55 months to 54 months.
Theses changes may even be an underestimate of the changes in birth spacing since
this particular group appear to have had access to sex selection even before it became
widespread as shown by the unequal sex ratio for the 1972--1984 period for women with
two girls.

Not surprisingly given these effects of sex selection, the third spell for the 
best-educated women is also the case with the clearest reversal in the spacing pattern by 
sex composition; the birth intervals with two girls are consistently longer than with one
or two sons no matter the percentile used.
[highlight others]

% An important exception to the relatively short median intervals come from the better 
% educated women, especially in the absence of sons.
% The longest median interval was 48 months for the third spell for urban women with twelve 
% or more years of education who had no boys, an increase of almost 15 months over
% four decades.
% Similary, urban women with eight to eleven years of education and no sons had median
% birth intervals of around 40 months, an increase of about 10 months.
% Finally, the best educated women now have a median second birth interval of around 42
% months whether their first child was a boy or a girl.

% Despite the strong son preference in spacing and sex ratios, there is also some indication 
% that parents prefer a mix of boys and girls;
% women with two boys \emph{and} a girl consistently have the lowest parity progression 
% probability in the fourth spell across education levels.

Did the predictions of declining use of sex selection come true?
There is no clear evidence for or against a reversal in the use of sex selection, with 
some cases showing increases in sex ratios between the last two periods, others little change, 
and some a decline.
The best-educated women is again a good illustration.
The percent boys born to women with two girls continued to increase over the last two
periods, but the likelihood of a third birth declined and when the first child was a girl, 
the sex ratio for the second birth dropped slightly as did the probability of a second birth. 
There is, however, also cases where there is clearly no abatement in the increasing use of 
sex selection. 
The sex ratios in the absence of girls continued to increase for rural women with
one to seven years of education and the likelihood of an additional birth remained high.

@@

First, 
over the four decades, birth intervals have lengthen, whether measured by the
25th, 50th, or 75th percentile.
These increases are larger the higher the education of the mother, the higher the parity, 
and the higher the percentile measure.
